---
title: |
  | Semiparametric Covariance Estimation
  | for Sparse Functional Data
author:
  - Valentin Patilea$^\dagger$
  - Jeffrey S. Racine$^\ddagger$
institute:
  - Univ Rennes, Ensai, CREST--UMR 9194$^\dagger$
  - McMaster University$^\ddagger$
date: 06/24/2026
bibliography: ../references.bib
link-citations: true
knitr:
  opts_chunk:
    autodep: true
    cache: true
    collapse: true
    echo: false
    fig.align: "center"
    message: false
    warning: false
format:
  revealjs:
    background-transition: fade
    center: true
    chalkboard: true
    css: custom.css
    embed-resources: false
    footer: "Sparse covariance functional data analysis | V. Patilea and J. S. Racine | jeffreyracine.github.io/spcovar-slides"
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          window.MathJax = { tex: { tags: 'ams' } };
          </script>
    incremental: true
    link-external-newwindow: true
    multiplex: true
    preview-links: auto
    self-contained-math: false
    show-notes: false
    slide-number: true
    theme: default
    touch: true
    transition: slide
editor: source
---

# Overview

- We propose a semiparametric covariance estimator for sparsely observed
  functional data.

- The estimator decomposes the covariance surface into:

  - nonparametric marginal scale functions;
  - a structured correlation surface.

- The correlation component is fit from a menu of positive-definite parametric
  candidate families.

- Model averaging uses data-driven weights to mitigate parametric model
  uncertainty.

- The resulting estimator is computationally attractive, retains flexibility
  where the data are most informative, and bridges fully parametric and fully
  nonparametric covariance estimation for sparse functional data analysis
  (FDA).

- All tuning parameters are selected from the data.

::: {.content-hidden when-profile="short"}
# Outline of Talk

- Sparse FDA covariance: what makes the problem hard

- A semiparametric decomposition

- Nonparametric mean and variance estimation with boundary correction

- Correlation families, model averaging, and implementation details

- Finite-sample evidence

- CD4 application and live covariance-surface demo

- Summary and next steps

::: {.notes}
As in the 2024 talk, do not march through proofs line by line. The audience
should understand the estimator, why it is useful, and what the evidence says.
Proofs will be added to the manuscript separately.
:::

:::

::: {.content-hidden when-profile="short"}
# Why Sparse FDA Covariance Is Hard

:::

## Classical Versus Functional Data

- Classical sample element: one observes a finite-dimensional draw $Y_i$ from a
  random vector $Y$:
  $$
  Y_i \in \mathbb R^d .
  $$

- Latent functional sample element: $X_i$ is a trajectory, an element of an
  infinite-dimensional function space, drawn from a random function $X(\cdot)$,
  indexed by a continuum $t \in \mathcal T$:
  $$
  X_i = \{X_i(t): t \in \mathcal T\}.
  $$

- Sparse functional observation: one observes only a small, noisy subset:
  $$
  Y_{ij} = X_i(T_{ij}) + \varepsilon_{ij},
        \qquad j=1,\ldots,M_i .
  $$

::: {.notes}
Begin with the sample element before proceeding. Here "infinite-dimensional"
means that the latent curve is an element of a function space such as
$L^2[0,1]$ or $C[0,1]$, not a vector in $\mathbb R^d$. In principle it can be
represented by infinitely many basis coefficients, while sparse FDA observes
only finitely many noisy point evaluations $Y_{ij}$.
:::

## Why Covariance?

- The covariance surface $C(s,t)=\operatorname{Cov}\{X(s),X(t)\}$ describes
  how latent process values move together across time.

- It is the object behind:

    - dependence along the continuum;
    - separation of latent signal from measurement noise;
    - dimension reduction through functional principal component analysis
      (FPCA)/eigenfunctions.

- For sparse noisy observations,
  $$
  \operatorname{Cov}(Y_{ij},Y_{ik})
  =
  C(T_{ij},T_{ik})+\sigma_\varepsilon^2 1\{j=k\}.
  $$

- Estimating $C(s,t)$ is the path from scattered measurements to latent
  trajectory structure.

::: {.notes}
Equivalent phrases to use verbally:

- the latent covariance surface $C(s,t)$;
- the covariance function of the latent process;
- $C(s,t)=\operatorname{Cov}\{X(s),X(t)\}$, the latent process covariance.
:::

## Example (Children's Growth)

::: {.callout-note appearance="simple"}

- Let $X_i(t)$ be child $i$'s latent growth trajectory at age $t$.

- The covariance surface tracks how growth at two ages varies together across
  children:
  $$
  C(s,t)=\operatorname{Cov}\{X(s),X(t)\}.
  $$

- If $C(2,12)$ is large and positive, children above average at age 2 tend also
  to be above average at age 12.

- This is why covariance is central:

    - it separates latent growth variation from measurement noise;
    - it reveals dependence across age;
    - it supports FPCA/eigenfunctions that summarize dominant growth patterns.

- In this example, principal components may correspond to overall size, timing
  of growth spurts, or other shape features (our focus here is strictly on
  covariance estimation).

:::

::: {.notes}
In this example, $X_i(t)$ is a conceptual latent growth trajectory, so its
units are determined by the growth outcome being modeled. If the outcome is
height, then $t$ is age, for example in years, and $X_i(t)$ is latent height,
for example in centimeters. The observed measurement
$Y_{ij}=X_i(T_{ij})+\varepsilon_{ij}$ is then also in centimeters, and the
measurement error $\varepsilon_{ij}$ has the same units.
:::

## Sparse Curves

- In dense FDA, the classical workflow is to smooth each subject's observations
  into a curve, then estimate population features such as the mean and
  covariance from the recovered curves.

- In sparse FDA, individual curve reconstruction is unstable or impossible, as
  the available covariance information is distributed across many subjects.

- The object of interest is the latent covariance function:
  $$
  \operatorname{Cov}\{X(s),X(t)\}.
  $$

- This object is learned from irregularly placed within-subject pairs.

- Replication matters.

## Measurement Error

- Let
  $$
  C(s,t)=\operatorname{Cov}\{X(s),X(t)\}
  $$
  denote the covariance function of the latent process $X$.

- The covariance of observations is not just this latent covariance:
  $$
  \operatorname{Cov}(Y_{ij},Y_{ik})
  = C(T_{ij},T_{ik}) + \sigma_\varepsilon^2 1\{j=k\}.
  $$

- The diagonal mixes latent variance and noise variance.

- The off-diagonal is informative but can be sparse and unevenly distributed.

::: {.notes}
The left-hand side,
$\operatorname{Cov}(Y_{ij},Y_{ik})$,
is the covariance of the observed noisy measurements, not the latent covariance function itself. Writing it explicitly makes the distinction clear:

observed covariance = latent covariance + measurement-error contribution
:::

## Why Direct Surface Smoothing Can Struggle

- Fully nonparametric covariance estimation targets a two-dimensional surface:
  $$
  C(s,t), \qquad (s,t)\in\mathcal T^2 .
  $$

- That is flexible, but sparse designs can leave large regions weakly informed.

- The estimator must manage:

    - irregular design points,
    - measurement error,
    - positive-definiteness pressure,
    - edge effects,
    - computation.

::: {.content-hidden when-profile="short"}
# Proposed Semiparametric Decomposition

:::

## The Object

- Let
  $$
  \mu(t)=\mathbb{E}\{X(t)\},
  \qquad
  \sigma^2(t)=\operatorname{Var}\{X(t)\},
  $$
  and

  $$
  \rho(s,t)=\operatorname{Corr}\{X(s),X(t)\}.
  $$

- Then
  $$
  C(s,t)=\sigma(s)\sigma(t)\rho(s,t).
  $$

## The Main Idea

- Estimate marginal scale nonparametrically.

- Fit the correlation component using low-dimensional positive-definite
  parametric families.

- Average the resulting covariance candidates with data-driven weights.

## Why This Is Useful

- This decomposition turns a difficult surface estimation problem into:

    1. two one-dimensional nonparametric regressions;
    2. one correlation-family fitting problem;
    3. a model-averaging problem over candidate correlation families.

- The computational burden is closer to parametric fitting than to unrestricted
  two-dimensional smoothing.

## Related Work

- Sparse FDA and FPCA foundations:

    - Principal Analysis by Conditional Expectation (PACE) estimates
      mean/covariance surfaces and recovers scores by conditional expectation
      [@Yao2005Sparse].
    - Functional snippets and `mcfda` use Fourier and semiparametric routes for
      mean/covariance estimation under severe sparse designs [@linwang2022].

- Direct covariance-surface estimation:

    - Fast Covariance Estimation (FACE) uses fast penalized-spline covariance
      smoothing [@Xiao2018SparseFACE].
    - RKHS methods regularize the covariance operator while enforcing positive
      semidefiniteness [@WongZhang2019].
    - Our `NP` benchmark uses a flexible tensor cosine covariance estimator.

- Semiparametric covariance modeling:

    - Variance-correlation decompositions motivate separating marginal scale
      from dependence structure [@Fan2007; @Fan2008].

::: {.notes}
PACE: Principal Analysis by Conditional Expectation.

FPCA: Functional Principal Component Analysis.

FACE: Fast Covariance Estimation.

RKHS: Reproducing Kernel Hilbert Space.

mcfda: Mean and Covariance Estimation for Functional Snippets and Sparse
Functional Data.
:::

## Assumptions

- Sparse observation model:
  $$
  Y_{ij}=X_i(T_{ij})+\varepsilon_{ij},
  \qquad
  1\le i\le N,\quad 1\le j\le M_i.
  $$

- $X_i(\cdot)$ are independent copies of a second-order latent process $X$ on
  $[0,1]$.

- The design points $T_{ij}$ are i.i.d. as $T$, and the errors
  $\varepsilon_{ij}$ are centered i.i.d. with variance
  $\sigma_\varepsilon^2$.

- The collections $\{X_i\}$, $\{T_{ij}\}$, and $\{\varepsilon_{ij}\}$ are
  mutually independent.

- The $M_i$ are bounded and treated as fixed; if random and independent, the
  analysis is conditional on the $M_i$.

::: {.content-visible when-profile="short"}
## Implementation Details 

- The default fit is data-driven rather than hand-tuned.

- The main tuning choices are selected from the data:

    - noise variance $\sigma_\varepsilon^2$: nearest within-subject contrasts
      with $k_{nn}\asymp \log n$;
    - mean and variance functions $\mu(t),\sigma^2(t)$: projection
      coefficients are estimated in an orthonormal cosine-type series basis,
      and a smoothed Efromovich threshold selects the retained cosine terms;
    - correlation families $\rho_\theta(s,t)$: low-dimensional parameters
      $\theta$ are fit from off-diagonal products, avoiding direct
      measurement-error contamination;
    - model averaging weights $\widehat{\boldsymbol{\alpha}}$: simplex weights are chosen by the
      same off-diagonal fitting criterion.

:::

::: {.content-hidden when-profile="short"}
## Implementation Details 

- The default fit is data-driven rather than hand-tuned.

- The main tuning choices are selected from the data:

    - noise variance $(\sigma_\varepsilon^2)$: nearest within-subject contrasts
      with $k_{nn}\asymp \log n$;
    - mean and variance functions $(\mu(t),\sigma^2(t))$: projection
      coefficients are estimated in an orthonormal boundary-corrected
      cosine-type basis; the boundary columns are retained, and a smoothed
      Efromovich threshold selects and shrinks the cosine tail;
    - nonparametric covariance benchmark $(C(s,t))$: tensor order $J$ and
      roughness penalty $\lambda$ are selected jointly by curve-level
      cross-validation or by a NOMAD search over $(J,\log_{10}\lambda)$;
    - correlation families $(\rho_\theta(s,t))$: low-dimensional parameters
      $\theta$ are fit from off-diagonal products, avoiding direct
      measurement-error contamination;
    - model averaging ($\widehat{\boldsymbol{\alpha}})$: simplex weights are chosen by the
      same off-diagonal fitting criterion.

:::

::: {.content-hidden when-profile="short"}
## A Statistical Compromise

- Fully parametric covariance:
  $$
  C_\theta(s,t)
  $$
  can be too rigid.

- Fully nonparametric covariance:
  $$
  \widehat C_{\mathrm{np}}(s,t)
  $$
  can be too variable under sparse designs.

- The proposed estimator lives between these two extremes.

:::

::: {.content-hidden when-profile="short"}
## Relation To Sparse FDA

- The sparse FDA setting follows the measurement-error and random-design
  framework used in the PACE literature [@Yao2005Sparse].

- Our contribution is not curve recovery itself.

- The focus is covariance estimation when one wants:

    - interpretable correlation structure,
    - flexible marginal variance,
    - low computational cost,
    - a route for model averaging.

:::

::: {.content-hidden when-profile="short"}
# Nonparametric Mean and Variance

:::

::: {.content-hidden when-profile="short"}
## Mean Estimation

- The mean is estimated from pooled observations:
  $$
  \widehat\mu(t)
  =\sum_{k=0}^{K}\widehat\theta_k \phi_k(t),
  $$

- Let $n=\sum_{i=1}^N M_i$ and let
  $U_1\leq\cdots\leq U_n$ denote the ordered pooled observation times.

- The coefficients are estimated from the aligned pooled responses, and the
  basis is selected and truncated from the pooled sample.

- The series step is one-dimensional and computationally inexpensive.

:::

::: {.content-hidden when-profile="short"}
## Variance Estimation

- After estimating the mean and noise variance, the latent marginal variance is
  estimated through the variance pseudo-response
$$
Z_{ij}^{(v)}
=\{Y_{ij}-\mu(T_{ij})\}^2-\sigma_\varepsilon^2.
$$

- The target is
  $$
  \sigma^2(t)
  =\mathbb{E}\{Z_{ij}^{(v)}\mid T_{ij}=t\}.
  $$

- The feasible version replaces $\mu$ and $\sigma_\varepsilon^2$ by estimates.

- The noise variance is estimated using very local within-subject contrasts.

:::

## Noise Variance

- For close within-curve pairs, the latent curve is nearly unchanged, so
  half-squared local differences mostly reflect measurement noise:
$$
D_{ijk}
=\frac12
\left[
\{Y_{ij}-\widehat\mu(T_{ij})\}
-\{Y_{ik}-\widehat\mu(T_{ik})\}
\right]^2 .
$$

- Average the $k_{nn}$ within-curve pairs with the smallest time gaps
  $|T_{ij}-T_{ik}|$:
$$
\widehat\sigma_\varepsilon^2
=\frac{1}{k_{nn}}\sum_{(i,j,k)\in\mathcal N_{k_{nn}}}D_{ijk}.
$$

- The operational rule is deliberately simple:
  $$
  k_{nn} \asymp \log n,
  \qquad n=\sum_{i=1}^N M_i.
  $$

## Efromovich's Cosine Series Estimator

- Efromovich's cosine-series approach is attractive because the cosine basis is
  orthonormal on $[0,1]$ and supports thresholding rules [@efromovich1999].

- For $k\geq 0$,
  $$
  \phi_k(t)=
  \begin{cases}
  1, & k=0,\\
  \sqrt{2}\cos(k\pi t), & k\geq 1 .
  \end{cases}
  $$

- This basis is stable, fast, and surprisingly competitive in finite samples.

- For a one-dimensional target $m(t)=\mathbb E(Z\mid T=t)$,
  $$
  m_J(t)=\sum_{k=0}^J\theta_k\phi_k(t),
  \qquad
  \theta_k=\int_0^1m(t)\phi_k(t)\,dt .
  $$

- The retained lag $J$ can be selected by a threshold/truncation
  rule or by cross-validation.

## Cosine Series Coefficients 

- The coefficients are moments:
  $$
  \theta_k=
  \mathbb E\!\left\{\frac{Z\phi_k(T)}{f_T(T)}\right\}.
  $$

- With ordered pooled times $U_\ell$, local spacings estimate density:
  $$
  \frac{1}{n f_T(U_\ell)}
  \approx
  \frac{U_{\ell+s}-U_{\ell-s}}{2s},
  \qquad
  s=\left\lceil\frac{1+\log\log(n+20)}{2}\right\rceil .
  $$

- Hence
  $$
  \widehat\theta_k
  \approx
  \sum_{\ell=1}^n
  Z_\ell\phi_k(U_\ell)
  \frac{U_{\ell+s}-U_{\ell-s}}{2s}.
  $$

::: {.content-hidden when-profile="short"}
## The Boundary Issue

- The cosine basis imposes
  $$
  \phi_k'(0)=\phi_k'(1)=0 .
  $$

- Any finite cosine expansion therefore has zero derivative at both endpoints.

- That can be helpful when it is true.

- It is restrictive when the mean or variance has genuine endpoint slope.

:::

::: {.content-hidden when-profile="short"}
## Efromovich's Remedy

- Efromovich discusses polynomial correction as a way to handle boundary
  incompatibility [@efromovich1999, ch. 2].

- The basic idea is:

    1. represent low-order endpoint behavior with polynomial terms;
    2. let the cosine tail estimate the remaining smooth component.

- This preserves the spirit of the cosine estimator while relaxing the endpoint
  constraint.

:::

::: {.content-hidden when-profile="short"}
## Orthonormality Matters

- Adding polynomial columns directly breaks the simple orthonormal structure
  that motivates thresholded series estimation.

- The remedy used here is to orthonormalize the finite set of boundary columns
  and cosine columns.

- Numerically, this is done with a QR/Gram-Schmidt construction
  [@bjorck2015].

:::

::: {.content-hidden when-profile="short"}
## Boundary-Corrected Basis

- For a quadratic correction, start from
  $$
  \{1,\; t,\; t^2,\; \sqrt{2}\cos(\pi t),\ldots,
  \sqrt{2}\cos(L\pi t)\}.
  $$

- Then construct an orthonormal basis spanning the same finite space.

- Linear correction allows a common endpoint slope.

- Quadratic correction allows independent endpoint slopes.

:::

::: {.content-hidden when-profile="short"}
## Why This Is Honest

- The uncorrected cosine estimator is not naive.

- It is a strong estimator when endpoint slopes are truly zero and is closely
  related to modified Fourier ideas for nonperiodic functions.

- Boundary correction is chosen because it avoids forcing zero derivatives at
  the endpoints.

- The price can be variance.

:::

::: {.content-hidden when-profile="short"}
## Bias-Variance Tradeoff

- In finite samples, boundary correction can increase integrated MSE even while
  removing a visible endpoint distortion.

- This is not a contradiction.

- It is the familiar statistical tradeoff:
  $$
  \operatorname{MSE}(t)
  = \operatorname{Bias}^2(t)+\operatorname{Var}(t).
  $$

::: {.notes}
This wording mirrors the current manuscript stance: default boundary correction
is a defensible choice because it does not build in the zero-derivative prior.
:::

:::

::: {.content-hidden when-profile="short"}
# Correlation Families

:::

::: {.content-hidden when-profile="short"}
## Why Correlation Families?

- Once $\sigma(t)$ is estimated, the remaining object is a correlation surface.

- Correlation families encode shape:

    - smoothness,
    - local roughness,
    - bridge-like behavior,
    - varying scale,
    - long-range decay.

- They also help maintain positive-definite structure.

:::

::: {.content-hidden when-profile="short"}
## Correlation Family Provenance

- Examples include:

    - Matern [@Matern1986]
    - multifractional Brownian motion [@PeltierLevyVehel1995]
    - Brownian bridge [@Doob1949]
    - Gaussian-process families [@RasmussenWilliams2006; @Gibbs1997]
    - generalized Cauchy [@LimTeo2009GeneralizedCauchy]
    - generalized Wendland [@Wendland2005; @Bevilacqua2019GeneralizedWendland]

:::

## Parametric Correlation Families for $\rho(s,t)$ {.smaller}

::: nonincremental

The set of candidate families is deliberately heterogeneous; model-averaging is over shapes, not just
over parameter values.

| Family | Source | Main feature |
|---|---|---|
| Matérn | @Matern1986 | stationary range and smoothness benchmark |
| mfBm | @PeltierLevyVehel1995 | time-varying roughness |
| Brownian bridge | @Doob1949 | endpoint-constrained nonstationarity |
| Rational quadratic | @RasmussenWilliams2006 | scale mixtures of local ranges |
| Gibbs | @Gibbs1997; @RasmussenWilliams2006 | time-varying local length scale |
| Generalized Cauchy | @LimTeo2009GeneralizedCauchy | separates local behavior and tail decay |
| Generalized Wendland | @Wendland2005; @Bevilacqua2019GeneralizedWendland | compact support with smoothness control |

:::

## Off-Diagonal Fitting

- For subject $i$, form centered products
  $$
  C_{ijk}
  =\{Y_{ij}-\widehat\mu(T_{ij})\}
  \{Y_{ik}-\widehat\mu(T_{ik})\},
  \qquad j\neq k.
  $$

- The off-diagonal avoids direct contamination by measurement error.

- Candidate correlations are fit to standardized products.

- For each family $\ell$, estimate
  $$
  \widehat\theta_\ell
  =
  \arg\min_{\theta_\ell}\left\{
  \frac{1}{N}\sum_{i=1}^N
  \frac{1}{M_i(M_i-1)}
  \sum_{j \ne k}
  \left[
  \widehat\sigma(T_{ij})\widehat\sigma(T_{ik})
  \rho_{\theta_\ell}^{(\ell)}(T_{ij},T_{ik})
  - C_{ijk}
  \right]^2\right\}.
  $$

## Model Averaging

- Consider one fitted covariance candidate per family,
$$
\widehat C_\ell(s,t)
=
\widehat\sigma(s)\widehat\sigma(t)\rho_{\widehat\theta_\ell}^{(\ell)}(s,t),
$$
where $\ell=1,\ldots,L$ indexes the candidate family.


- Instead of selecting one correlation family, compute weights
  $$
  \arg\min_{\boldsymbol{\alpha}\in\Delta}\left\{
  \frac{1}{N}\sum_{i=1}^N
  \frac{1}{M_i(M_i-1)}
  \sum_{j \ne k}
  \left[
  \sum_{\ell=1}^L \alpha_\ell
  \widehat\sigma(T_{ij})\widehat\sigma(T_{ik})
  \rho_{\widehat\theta_\ell}^{(\ell)}(T_{ij},T_{ik})
  - C_{ijk}
  \right]^2\right\}.
  $$

- The averaged estimator is
  $\widehat C_{\boldsymbol{\alpha}}(s,t)=\sum_{\ell=1}^L
  \alpha_\ell \widehat C_\ell(s,t)$, with $\alpha_\ell \ge 0$ and
  $\sum_{\ell=1}^L \alpha_\ell = 1$.


- This follows the spirit of focused/model-averaging ideas
  [@CLAESKENS_HJORT:2011].

::: {.content-hidden when-profile="short"}
## Why Average?

- Hard selection is brittle when families are close.

- Model averaging allows:

    - dominant weight on a good family,
    - insurance when two families explain different regions,
    - interpretable diagnostics via $\widehat\boldsymbol{\alpha}$.

- The averaging weights are part of the inferential output.

:::

::: {.content-hidden when-profile="short"}
# Parametric and Nonparametric Benchmarks

:::

::: {.content-hidden when-profile="short"}
## Fully Parametric Benchmark

- The fully parametric benchmark estimates:
  $$
  \mu_\eta(t),\qquad
  \sigma^2_\gamma(t),\qquad
  \rho_\theta(s,t).
  $$

- This is stable and fast when the structure is right.

- It can be misleading when location, scale, or correlation is misspecified.

:::

::: {.content-hidden when-profile="short"}
## Fully Nonparametric Benchmark

- The nonparametric covariance benchmark estimates the surface directly:
  $$
  C(s,t).
  $$

- This is flexible, but the tensor problem is data-hungry under sparse sampling.

- It is a useful competitor and a useful diagnostic.

:::

::: {.content-hidden when-profile="short"}
## Combined-Test Diagnostic

- The manuscript also compares convex combinations such as
  $$
  \lambda\widehat C_{\mathrm{param}}
  +(1-\lambda)\widehat C_{\mathrm{sp}}.
  $$

- If the optimum places weight near one estimator, that is informative.

- If it mixes, that is also informative.

:::

::: {.content-hidden when-profile="short"}
# Finite-Sample Evidence

```{r peer-ratio-table-setup}
#| include: false
library(knitr)

pretty_simulation <- function(x) {
  map <- c(
    "Matern" = "Matérn",
    "GeneralizedCauchy" = "Generalized Cauchy",
    "GeneralizedWendland" = "Generalized Wendland",
    "RationalQuadratic" = "Rational Quadratic"
  )
  ifelse(x %in% names(map), unname(map[x]), x)
}

pretty_name <- function(x) {
  map <- c(
    "spcovar default" = "SP",
    "spcovar default minus DGP family" = "SP-OMIT",
    "Matern" = "Matérn",
    "PACE/fdapace" = "PACE",
    "mcfda Fourier" = "mcfda-F",
    "mcfda SP" = "mcfda-SP",
    "GeneralizedCauchy" = "GC",
    "GeneralizedWendland" = "GW",
    "RationalQuadratic" = "RQ",
    "npcovar" = "NP",
    "simulation" = "DGP"
  )
  if (x %in% names(map)) map[[x]] else x
}

make_peer_ratio_table <- function(dat,
                                  denominator,
                                  include_methods,
                                  sim_order,
                                  regime = NULL,
                                  digits = 2L) {
  if (!nrow(dat)) return(NULL)
  if (!is.null(regime)) {
    dat <- dat[dat$regime == regime, , drop = FALSE]
  }
  if (!nrow(dat)) return(NULL)

  denom <- dat[dat$method_label == denominator,
               c("simulation", "regime", "mean_mse_x1000")]
  names(denom)[3L] <- "den_mse"
  numer <- dat[dat$method_label %in% include_methods,
               c("simulation", "regime", "method_label", "mean_mse_x1000")]
  names(numer)[4L] <- "num_mse"
  merged <- merge(numer, denom, by = c("simulation", "regime"))
  merged <- merged[
    is.finite(merged$num_mse) &
      is.finite(merged$den_mse) &
      merged$num_mse > 0 &
      merged$den_mse > 0,
    ,
    drop = FALSE
  ]
  if (!nrow(merged)) return(NULL)
  merged$ratio <- merged$num_mse / merged$den_mse

  make_cell <- function(x) {
    if (!length(x)) return("")
    formatC(exp(mean(log(x))), digits = digits, format = "f")
  }

  rows <- list()
  for (sim in sim_order) {
    row <- c(simulation = sim)
    for (m in include_methods) {
      row[[m]] <- make_cell(merged$ratio[
        merged$simulation == sim & merged$method_label == m
      ])
    }
    rows[[length(rows) + 1L]] <- as.data.frame(as.list(row), stringsAsFactors = FALSE)
  }

  overall <- c(simulation = "Overall")
  for (m in include_methods) {
    overall[[m]] <- make_cell(merged$ratio[merged$method_label == m])
  }
  rows[[length(rows) + 1L]] <- as.data.frame(as.list(overall), stringsAsFactors = FALSE)

  out <- do.call(rbind, rows)
  names(out) <- c("simulation", include_methods)
  rownames(out) <- NULL
  out
}

reorder_table_by_row_values <- function(tab,
                                        row_value = "Overall",
                                        leading_cols = 1L,
                                        decreasing = FALSE) {
  if (is.null(tab) || !nrow(tab) || ncol(tab) <= leading_cols) return(tab)
  row_idx <- which(tab[[1L]] == row_value)
  if (!length(row_idx)) return(tab)
  ordered_cols <- seq.int(leading_cols + 1L, ncol(tab))
  scores <- suppressWarnings(as.numeric(tab[row_idx[1L], ordered_cols, drop = TRUE]))
  tab[, c(seq_len(leading_cols), ordered_cols[order(scores, decreasing = decreasing, na.last = TRUE)]), drop = FALSE]
}

render_peer_ratio_table <- function(tab) {
  if (is.null(tab) || !nrow(tab)) {
    cat("Peer comparison summary is not available.")
    return(invisible(NULL))
  }
  kable(
    tab,
    format = "html",
    align = c("l", rep("r", ncol(tab) - 1L)),
    col.names = vapply(names(tab), pretty_name, character(1L), USE.NAMES = FALSE),
    row.names = FALSE,
    escape = FALSE,
    table.attr = 'class="peer-ratio-table"'
  )
}

load_peer_campaign_summaries <- function(campaign_dir, cfg_env) {
  config_path <- file.path(campaign_dir, "campaign_config.R")
  results_dir <- file.path(campaign_dir, "results_full")
  if (!file.exists(config_path) || !dir.exists(results_dir)) return(data.frame())

  out <- list()
  for (key in cfg_env$campaign_dgp_keys()) {
    cfg <- cfg_env$campaign_dgp_config(key)
    for (multiplier in cfg_env$campaign_regime_multipliers) {
      regime_slug <- tolower(cfg_env$campaign_regime_label(multiplier))
      summary_path <- file.path(
        results_dir,
        key,
        sprintf("%s_%s_summary.csv", cfg$prefix, regime_slug)
      )
      if (!file.exists(summary_path)) next
      dat <- read.csv(summary_path, stringsAsFactors = FALSE)
      if (!nrow(dat)) next
      dat$simulation <- cfg$short_label
      dat$method_label <- unname(cfg_env$campaign_method_labels[dat$method])
      out[[length(out) + 1L]] <- dat
    }
  }
  if (!length(out)) data.frame() else do.call(rbind, out)
}

load_peer_campaign_alphas <- function(campaign_dir, cfg_env) {
  config_path <- file.path(campaign_dir, "campaign_config.R")
  results_dir <- file.path(campaign_dir, "results_full")
  if (!file.exists(config_path) || !dir.exists(results_dir)) return(data.frame())

  out <- list()
  for (key in cfg_env$campaign_dgp_keys()) {
    cfg <- cfg_env$campaign_dgp_config(key)
    for (multiplier in cfg_env$campaign_regime_multipliers) {
      regime_slug <- tolower(cfg_env$campaign_regime_label(multiplier))
      alpha_path <- file.path(
        results_dir,
        key,
        sprintf("%s_%s_alpha_summary.csv", cfg$prefix, regime_slug)
      )
      if (!file.exists(alpha_path)) next
      dat <- read.csv(alpha_path, stringsAsFactors = FALSE)
      if (!nrow(dat)) next
      dat$simulation <- cfg$short_label
      dat$method_label <- unname(cfg_env$campaign_method_labels[dat$method])
      out[[length(out) + 1L]] <- dat
    }
  }
  if (!length(out)) data.frame() else do.call(rbind, out)
}

build_peer_ratio_tables <- function(campaign_dir) {
  campaign_dir <- normalizePath(campaign_dir, mustWork = FALSE)
  config_path <- file.path(campaign_dir, "campaign_config.R")
  if (!file.exists(config_path)) {
    return(list(overall = NULL, N = NULL, `2N` = NULL, `4N` = NULL))
  }

  cfg_env <- new.env(parent = globalenv())
  source(config_path, local = cfg_env)

  summary_dat <- load_peer_campaign_summaries(campaign_dir, cfg_env)
  if (!nrow(summary_dat)) {
    return(list(overall = NULL, N = NULL, `2N` = NULL, `4N` = NULL))
  }

  summary_dat$simulation <- pretty_simulation(summary_dat$simulation)
  sim_order <- sort(pretty_simulation(vapply(
    cfg_env$campaign_dgp_keys(),
    function(key) cfg_env$campaign_dgp_config(key)$short_label,
    character(1L)
  )))
  peer_methods <- unname(cfg_env$campaign_method_labels[c(
    "npcovar_nomad",
    "face_sparse",
    "fdapace",
    "rkhscovfun_cv",
    "mcfda_fourier",
    "mcfda_sp"
  )])
  sp_methods <- unname(cfg_env$campaign_method_labels[c(
    "spcovar_default",
    "spcovar_default_minus_own_family"
  )])
  table_overall <- make_peer_ratio_table(
    summary_dat,
    denominator = sp_methods[1L],
    include_methods = c(sp_methods[2L], peer_methods),
    sim_order = sim_order
  )
  table_overall <- reorder_table_by_row_values(table_overall)
  ordered_methods <- names(table_overall)[-1L]
  table_N <- make_peer_ratio_table(
    summary_dat,
    denominator = sp_methods[1L],
    include_methods = ordered_methods,
    sim_order = sim_order,
    regime = "N"
  )
  table_2N <- make_peer_ratio_table(
    summary_dat,
    denominator = sp_methods[1L],
    include_methods = ordered_methods,
    sim_order = sim_order,
    regime = "2N"
  )
  table_4N <- make_peer_ratio_table(
    summary_dat,
    denominator = sp_methods[1L],
    include_methods = ordered_methods,
    sim_order = sim_order,
    regime = "4N"
  )
  list(overall = table_overall, N = table_N, `2N` = table_2N, `4N` = table_4N)
}

make_alpha_table <- function(dat,
                             method_label,
                             families,
                             sim_order,
                             regime = NULL,
                             digits = 2L) {
  if (!nrow(dat)) return(NULL)
  sub <- dat[dat$method_label == method_label, , drop = FALSE]
  if (!is.null(regime)) {
    sub <- sub[sub$regime == regime, , drop = FALSE]
  }
  if (!nrow(sub)) return(NULL)

  family_value <- function(x, fam) {
    if (!fam %in% names(x) || is.na(x[[fam]])) 0 else x[[fam]]
  }

  rows <- list()
  for (sim in sim_order) {
    ss <- sub[sub$simulation == sim, , drop = FALSE]
    if (!nrow(ss)) next
    means <- tapply(ss$alpha, ss$family, mean)
    row <- c(DGP = sim)
    for (fam in families) {
      row[[fam]] <- formatC(family_value(means, fam), digits = digits, format = "f")
    }
    rows[[length(rows) + 1L]] <- as.data.frame(as.list(row), stringsAsFactors = FALSE)
  }

  overall_means <- tapply(sub$alpha, sub$family, mean)
  overall_row <- c(DGP = "Overall")
  for (fam in families) {
    overall_row[[fam]] <- formatC(family_value(overall_means, fam), digits = digits, format = "f")
  }
  rows[[length(rows) + 1L]] <- as.data.frame(as.list(overall_row), stringsAsFactors = FALSE)

  out <- do.call(rbind, rows)
  names(out) <- c("DGP", families)
  rownames(out) <- NULL
  out
}

build_peer_alpha_tables <- function(campaign_dir) {
  campaign_dir <- normalizePath(campaign_dir, mustWork = FALSE)
  config_path <- file.path(campaign_dir, "campaign_config.R")
  if (!file.exists(config_path)) {
    empty <- list(overall = NULL, N = NULL, `2N` = NULL, `4N` = NULL)
    return(list(default = empty, drop = empty))
  }

  cfg_env <- new.env(parent = globalenv())
  source(config_path, local = cfg_env)

  alpha_dat <- load_peer_campaign_alphas(campaign_dir, cfg_env)
  if (!nrow(alpha_dat)) {
    empty <- list(overall = NULL, N = NULL, `2N` = NULL, `4N` = NULL)
    return(list(default = empty, drop = empty))
  }

  alpha_dat$simulation <- pretty_simulation(alpha_dat$simulation)
  sim_order <- sort(pretty_simulation(vapply(
    cfg_env$campaign_dgp_keys(),
    function(key) cfg_env$campaign_dgp_config(key)$short_label,
    character(1L)
  )))
  families <- cfg_env$campaign_default_families[
    order(pretty_simulation(cfg_env$campaign_default_families))
  ]
  sp_methods <- unname(cfg_env$campaign_method_labels[c(
    "spcovar_default",
    "spcovar_default_minus_own_family"
  )])

  build_for_method <- function(method_label) {
    list(
      overall = make_alpha_table(alpha_dat, method_label, families, sim_order),
      N = make_alpha_table(alpha_dat, method_label, families, sim_order, regime = "N"),
      `2N` = make_alpha_table(alpha_dat, method_label, families, sim_order, regime = "2N"),
      `4N` = make_alpha_table(alpha_dat, method_label, families, sim_order, regime = "4N")
    )
  }

  list(
    default = build_for_method(sp_methods[1L]),
    drop = build_for_method(sp_methods[2L])
  )
}

render_alpha_table <- function(tab) {
  if (is.null(tab) || !nrow(tab)) {
    cat("Averaging-weight summary is not available.")
    return(invisible(NULL))
  }
  kable(
    tab,
    format = "html",
    align = c("l", rep("r", ncol(tab) - 1L)),
    col.names = vapply(names(tab), pretty_name, character(1L), USE.NAMES = FALSE),
    row.names = FALSE,
    escape = FALSE,
    table.attr = 'class="peer-ratio-table"'
  )
}

bc_peer_tables <- build_peer_ratio_tables("../Work_bcgs2_peer_beta_mc")
nbc_peer_tables <- build_peer_ratio_tables("../Work_nbc_peer_uniform_mc")
bc_alpha_tables <- build_peer_alpha_tables("../Work_bcgs2_peer_beta_mc")
```

:::

## Simulation Design

- The revised simulation design emphasizes:

    - beta-mixture observation times;
    - sparse within-subject sampling;
    - non-Neumann mean and variance functions;
    - seven correlation DGPs in the peer comparison;
    - 1000 replications for final tables.

- This is intentionally harder than a uniform-design, zero-endpoint-slope
  setting.

::: {.content-hidden when-profile="short"}
## Peer Comparison

- The peer comparison places the proposed estimator against available R
  implementations for sparse functional covariance estimation.

- The point is practical:

    - what happens with the same data-generating designs?
    - which methods return complete estimates?
    - how do MSE ratios and rankings change across DGPs?

:::

## Peer Estimators In The Simulation {.smaller}

::: nonincremental

The peer comparison uses available R implementations where possible; the table
labels refer to the implemented estimators.

| Label | Source / implementation | Core strategy | Main contrast with SP |
|---|---|---|---|
| SP | proposed `spcovar` | nonparametric margins, parametric correlation library, convex averaging | low-dimensional bivariate structure |
| SP-OMIT | proposed sensitivity fit | SP with the true DGP family removed | robustness check, not an external peer |
| NP | proposed `npcovar` | tensor cosine covariance with data-driven tuning | flexible surface, less structure |
| PACE | `fdapace::FPCA()` [@Yao2005Sparse; @fdapacePackage2024] | sparse FPCA via smoothed covariance and conditional scores | dimension reduction rather than correlation averaging |
| FACE | `face::face.sparse()` [@Xiao2018SparseFACE; @facePackage2025] | fast penalized-spline covariance smoothing | direct covariance smoothing |
| RKHS | `rkhscovfun::rkhscov.cv()` [@WongZhang2019; @rkhscovfunPackage2018] | positive semidefinite RKHS covariance regularization | flexible PSD nonparametric covariance |
| mcfda-F | `mcfda` Fourier route [@linwang2022; @mcfdaPackage] | Fourier mean and covariance estimation | Fourier covariance route |
| mcfda-SP | `mcfda` SP route [@linwang2022; @mcfdaPackage] | semiparametric covariance route in `mcfda` | external SP-like implementation |

:::

::: {.notes}
Exact campaign calls:

- `SP`: `spcovar(..., mean_lag_method = "smoothed", var_lag_method = "smoothed")`.
- `SP-OMIT`: same call with `fit_families = campaign_drop_own_family(dgp_key)`.
- `NP`: `npcovar(..., nomad = TRUE, standardize = TRUE)`.
- `PACE`: `fdapace::FPCA(..., dataType = "Sparse", error = TRUE, methodXi = "CE", maxK = 5, methodSelectK = "FVE", FVEthreshold = 0.99)`.
- `FACE`: `face::face.sparse(..., calculate.scores = FALSE, search.length = 8)`.
- `RKHS`: `rkhscovfun::rkhscov.cv(..., nfold = 3, pos = TRUE, control.cv = list(nlam = 6, gams = 1))`.
- `mcfda-F`: `mcfda::meanfunc(..., method = "FOURIER", tuning = "cv")` followed by `mcfda::covfunc(..., method = "FOURIER", tuning = "lle")`.
- `mcfda-SP`: `mcfda::covfunc(..., method = "SP", weig = NULL)`.
:::

## Peer Comparison Ratios {.table-slide}

::: nonincremental

Each entry is $\mathrm{MSE}(\text{method})/\mathrm{MSE}(\text{proposed})$;
smaller is better. The first tab is the manuscript's overall geometric index,
and the remaining tabs fix the sparse-sample regime.

Beta-time campaign ($M_i\in\{1,\ldots,11\}$, baseline $N\approx42$, so
$\mathbb E[\sum_i M_i]\approx252$).

::: {.panel-tabset .peer-tabs}

### Overall

```{r}
#| echo: false
render_peer_ratio_table(bc_peer_tables$overall)
```

### N

```{r}
#| echo: false
render_peer_ratio_table(bc_peer_tables$N)
```

### 2N

```{r}
#| echo: false
render_peer_ratio_table(bc_peer_tables$`2N`)
```

### 4N

```{r}
#| echo: false
render_peer_ratio_table(bc_peer_tables$`4N`)
```

:::

:::

::: {.content-hidden when-profile="short"}
## Peer Comparison Ratios: Not Bias Corrected {.table-slide}

::: nonincremental

Each entry is $\mathrm{MSE}(\text{method})/\mathrm{MSE}(\text{proposed})$;
smaller is better. The first tab is the overall geometric index, and the
remaining tabs fix the sparse-sample regime.

No-boundary-correction uniform-time control campaign
($M_i\in\{1,\ldots,11\}$, baseline $N\approx42$, so
$\mathbb E[\sum_i M_i]\approx252$).

::: {.panel-tabset .peer-tabs}

### Overall

```{r}
#| echo: false
render_peer_ratio_table(nbc_peer_tables$overall)
```

### N

```{r}
#| echo: false
render_peer_ratio_table(nbc_peer_tables$N)
```

### 2N

```{r}
#| echo: false
render_peer_ratio_table(nbc_peer_tables$`2N`)
```

### 4N

```{r}
#| echo: false
render_peer_ratio_table(nbc_peer_tables$`4N`)
```

:::

:::

:::

::: {.content-hidden when-profile="short"}
## Averaging Weights

- The $\widehat{\boldsymbol{\alpha}}$ weights are diagnostic, not just
  computational detail.

- They indicate which correlation families explain the simulated or empirical
  surface.

- The manuscript tables report both:

    - weights across DGPs;
    - weights when the generating family is removed.

:::

## Averaging Weights: Proposed {.table-slide}

::: nonincremental

Average $\widehat{\boldsymbol{\alpha}}$ weights for the proposed estimator. The first tab
averages over all available DGP-regime cells; the remaining tabs fix the
sparse-sample regime.

Beta-time campaign ($M_i\in\{1,\ldots,11\}$, baseline $N\approx42$, so
$\mathbb E[\sum_i M_i]\approx252$).

::: {.panel-tabset .peer-tabs}

### Overall

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$default$overall)
```

### N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$default$N)
```

### 2N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$default$`2N`)
```

### 4N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$default$`4N`)
```

:::

:::

## Averaging Weights: Family Omitted {.table-slide}

::: nonincremental

Average $\widehat{\boldsymbol{\alpha}}$ weights when the DGP family is removed from the
candidate menu. The zero diagonal marks the omitted generating family.

Beta-time campaign ($M_i\in\{1,\ldots,11\}$, baseline $N\approx42$, so
$\mathbb E[\sum_i M_i]\approx252$).

::: {.panel-tabset .peer-tabs}

### Overall

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$drop$overall)
```

### N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$drop$N)
```

### 2N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$drop$`2N`)
```

### 4N

```{r}
#| echo: false
render_alpha_table(bc_alpha_tables$drop$`4N`)
```

::: {.notes}
The tables are generated from campaign summaries so they remain synchronized
with the manuscript evidence.
:::

:::

:::

::: {.content-hidden when-profile="short"}
# CD4 Application

:::

## Example (CD4 Data)

::: {.callout-note appearance="simple"}

- The application uses CD4 count data distributed with `refund` [@refund2026].

- CD4 counts are recorded relative to HIV seroconversion.

- Higher CD4 counts generally indicate stronger immune function.

- Lower CD4 counts are associated with greater immune damage.

- The sample contains 366 individuals and 1,888 pooled observations.

- Observation times are irregular, sparse, and normalized to $[0,1]$ before
  fitting.

::: {.fragment .cd4-count-fragment}

```{r}
#| echo: false
data(cd4, package = "refund")
point_count_tab_slide <- table(rowSums(is.finite(cd4)))
point_count_display_slide <- rbind(
  `Points per curve` = names(point_count_tab_slide),
  `Number of curves` = as.character(as.integer(point_count_tab_slide))
)
knitr::kable(
  point_count_display_slide,
  format = "html",
  align = c("l", rep("r", ncol(point_count_display_slide) - 1L)),
  row.names = TRUE,
  col.names = NULL,
  table.attr = 'class="cd4-count-table"'
)
```

:::

:::

::: {.notes}
CD4 refers to CD4-positive T lymphocytes, a class of immune cells targeted by
HIV. CD4 cell counts are a standard clinical marker of immune status in HIV
infection: lower counts indicate immune suppression and greater disease
progression risk. Seroconversion is the transition point at which HIV antibodies
become detectable, so measuring time relative to seroconversion gives a common
clinical origin for the longitudinal trajectories.
:::

## Example (CD4 - Why This Example Matters)

::: {.callout-note appearance="simple"}

- Seroconversion anchors follow-up to a biologically meaningful event.

- The scientific question is not only the average CD4 trajectory.

- We also need to understand:

    - how between-patient variability changes over time;
    - how strongly early and later immune status are associated;
    - how much variation is latent signal rather than measurement noise.

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - What The Estimator Must Separate)

::: {.callout-note appearance="simple"}

- The CD4 example is a canonical sparse longitudinal FDA setting.

- The observed trajectories are irregular and noisy.

- A covariance estimator must separate:

    - population mean trajectory,
    - marginal variance over time,
    - within-subject correlation,
    - measurement noise.

:::

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Parametric Benchmark In The Application)

::: {.callout-note appearance="simple"}

- A simple parametric benchmark is feasible and interpretable.

- But it imposes strong structure:

    - linear mean;
    - constant latent variance;
    - Gaussian-shaped stationary covariance ridge;
    - no sharp separation of latent variation and measurement error.

:::

:::

```{r}
#| label: cd4-default-fits
#| include: false

if (!requireNamespace("spcovar", quietly = TRUE)) {
  stop("The spcovar package is required to render the CD4 slides.", call. = FALSE)
}
if (!requireNamespace("refund", quietly = TRUE)) {
  stop("The refund package is required to render the CD4 slides.", call. = FALSE)
}

library(spcovar)

data(cd4, package = "refund")

times <- as.numeric(colnames(cd4))
times01 <- (times - min(times)) / diff(range(times))

Lt <- vector("list", nrow(cd4))
Ly <- vector("list", nrow(cd4))

for (i in seq_len(nrow(cd4))) {
  ok <- is.finite(cd4[i, ])
  Lt[[i]] <- times01[ok]
  Ly[[i]] <- as.numeric(cd4[i, ok])
}

fit_cd4_sp <- spcovar(Lt, Ly ,
                      mean_basis = "cosine_bc_gs2",
                      var_basis = "cosine_bc_gs2")
fit_cd4_np <- npcovar(Lt, Ly,
                      nomad = TRUE,
                      standardize = TRUE,
                      mean_basis = "cosine_bc_gs2",
                      var_basis = "cosine_bc_gs2",
                      covariance_basis = "cosine_bc_gs2")
```

## Example (CD4 - Default Semiparametric Fit)

::: nonincremental

```{r}
#| label: fig-cd4-spcovar-default
#| fig-width: 8.5
#| fig-height: 5.0
#| fig.align: "center"
plot(fit_cd4_sp)
```

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Default Nonparametric Fit)

::: nonincremental

```{r}
#| label: fig-cd4-npcovar-default
#| fig-width: 8.5
#| fig-height: 5.0
#| fig.align: "center"
plot(fit_cd4_np)
```

:::

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Semiparametric Covariance Contour View) {#covariance-contour-views}

::: nonincremental

Heat maps and contours are a popular way to visualize covariance estimates.

```{r}
#| label: fig-cd4-spcovar-heatmap-default
#| fig-width: 7.8
#| fig-height: 5.2
#| fig.align: "center"
plot(fit_cd4_sp, view = "heatmap")
```

:::

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Nonparametric Covariance Contour View)

::: nonincremental

```{r}
#| label: fig-cd4-npcovar-heatmap-default
#| fig-width: 7.8
#| fig-height: 5.2
#| fig.align: "center"
plot(fit_cd4_np, view = "heatmap")
```

:::

:::

## Example (CD4 - Semiparametric Covariance Surface)

```{r}
#| label: fig-cd4-spcovar-surface-default
#| fig-width: 7.8
#| fig-height: 5.2
#| fig.align: "center"
plot(fit_cd4_sp, view = "persp")
```

::: {.content-hidden when-profile="short"}
## Example (CD4 - Nonparametric Covariance Surface)

::: nonincremental

```{r}
#| label: fig-cd4-npcovar-surface-default
#| fig-width: 7.8
#| fig-height: 5.2
#| fig.align: "center"
plot(fit_cd4_np, view = "persp")
```

:::

:::

::: {.content-visible when-profile="short"}

## Example (R Code)

::: {.callout-note appearance="simple"}
##  CD4 - Live rgl Demo Code using the R package `spcovar`

```{r}
#| echo: true
#| eval: false
#| label: fig-fexbwselcddens

## The R package spcovar is installed from GitHub until it is available on CRAN.
## The remotes package is the lightweight installer route.

if (!requireNamespace("spcovar", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
  }
  remotes::install_github("JeffreyRacine/spcovar")
}

required_packages <- c("refund", "rgl")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages)) {
  stop(
    "Please install the required package(s): ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

## Load the spcovar package, which contains the proposed estimator and plotting
## methods.

library(spcovar)

## Load the CD4 cell count data from the refund package.

data(cd4, package = "refund")

## The data are in a matrix with rows corresponding to subjects and columns
## corresponding to time points.  We will convert the time points to the unit
## interval and create lists of time points and observations for each subject,
## which is the format required by spcovar.

times <- as.numeric(colnames(cd4))
times01 <- (times - min(times)) / diff(range(times))

Lt <- vector("list", nrow(cd4))
Ly <- vector("list", nrow(cd4))

for (i in seq_len(nrow(cd4))) {
  ok <- is.finite(cd4[i, ])
  Lt[[i]] <- times01[ok]
  Ly[[i]] <- as.numeric(cd4[i, ok])
}

## Estimate the mean, standard deviation, and covariance surface using spcovar.
## See ?spcovar for options to customize the fit, including the mean and
## variance.

fit_cd4_sp <- spcovar(Lt = Lt, Ly = Ly)

## Plot the mean, standard deviation, and covariance surface estimates.
## See ?plot.spcovar_fit for options to customize the plot.

plot(fit_cd4_sp)

## Plot the covariance surface estimate in an interactive rgl widget.

plot(fit_cd4_sp, view = "rgl")
```

:::

:::

::: {.content-hidden when-profile="short"}

## Example (R Code)

::: {.callout-note appearance="simple"}
##  CD4 - Live rgl Demo Code using the R package `spcovar`

::: nonincremental

```{r}
#| echo: true
#| eval: false
#| label: fig-fexbwselcddens-long

## The R package spcovar is installed from GitHub until it is available on CRAN.
## The remotes package is the lightweight installer route.

if (!requireNamespace("spcovar", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
  }
  remotes::install_github("JeffreyRacine/spcovar")
}

## crs provides the NOMAD-backed tuning route used by the nonparametric fit.

required_packages <- c("refund", "rgl", "crs")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages)) {
  stop(
    "Please install the required package(s): ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

## Load the spcovar package, which contains the proposed estimator and plotting
## methods.

library(spcovar)

## Load the CD4 cell count data from the refund package.

data(cd4, package = "refund")

## The data are in a matrix with rows corresponding to subjects and columns
## corresponding to time points.  We will convert the time points to the unit
## interval and create lists of time points and observations for each subject,
## which is the format required by spcovar.

times <- as.numeric(colnames(cd4))
times01 <- (times - min(times)) / diff(range(times))

Lt <- vector("list", nrow(cd4))
Ly <- vector("list", nrow(cd4))

for (i in seq_len(nrow(cd4))) {
  ok <- is.finite(cd4[i, ])
  Lt[[i]] <- times01[ok]
  Ly[[i]] <- as.numeric(cd4[i, ok])
}

## Estimate the mean, standard deviation, and covariance surface using spcovar.
## See ?spcovar for options to customize the fit, including the mean and
## variance.

fit_cd4_sp <- spcovar(Lt = Lt, Ly = Ly)

## Plot the mean, standard deviation, and covariance surface estimates.
## See ?plot.spcovar_fit for options to customize the plot.

plot(fit_cd4_sp)

## Plot the covariance surface estimate in an interactive rgl widget.

plot(fit_cd4_sp, view = "rgl")

## Estimate the mean, standard deviation, and fully nonparametric covariance
## surface using npcovar. See ?npcovar for options to customize the fit,
## including the mean and variance.

fit_cd4_np <- npcovar(
  Lt = Lt,
  Ly = Ly,
  nomad = TRUE,
  standardize = TRUE
)

## Plot the mean, standard deviation, and covariance surface estimates.
## See ?plot.npcovar_fit for options to customize the plot.

plot(fit_cd4_np)

## Plot the fully nonparametric covariance surface estimate in an interactive
## rgl widget.

plot(fit_cd4_np, view = "rgl")
```

:::

:::

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Interpretation)

::: {.callout-note appearance="simple"}

- The application illustrates why boundary behavior matters in practice.

- Uncorrected cosine stages can visibly deform endpoint behavior by imposing
  zero derivatives.

- Boundary-corrected stages avoid that prior restriction and support sensitivity
  analysis across plausible choices.

:::

:::

::: {.content-hidden when-profile="short"}
## Example (CD4 - Sensitivity Is A Feature)

::: {.callout-note appearance="simple"}

- No boundary rule is uniformly best.

- The practical recommendation is to examine sensitivity across:

    - uncorrected cosine,
    - linear boundary correction,
    - quadratic boundary correction,
    - empirical QR alternatives.

- The default should be defensible, not magical.

:::

:::

::: {.content-hidden when-profile="short"}
# Summary

:::

## Main Points

- The proposed estimator:

    - decomposes covariance into scale and correlation;
    - estimates location and scale nonparametrically;
    - fits interpretable correlation families;
    - uses model averaging rather than hard family selection;
    - chooses the main tuning quantities from the data;
    - remains computationally attractive for sparse FDA;
    - avoids imposing zero endpoint derivatives by default.

::: {.content-hidden when-profile="short"}
## What Is Coming Next

- The current manuscript is intentionally ahead of the completed proof section.

- The theory will formalize:

    - consistency of the staged estimator;
    - behavior of the nonparametric mean and variance stages;
    - correlation-family fitting;
    - model-averaging behavior;
    - comparison with parametric and nonparametric benchmarks.

:::

## Closing Thoughts

- Sparse covariance estimation is a setting where structure helps.

- The goal is not to choose between parametric and nonparametric estimation.

- The goal is to put structure where it is useful, keep flexibility where it is
  needed, and make the resulting estimator practical.

::: {.content-hidden when-profile="short"}
# Backup Slides
:::

::: {.content-hidden when-profile="short"}
## Backup: Observation Model

- For independent subjects $i=1,\ldots,N$,
  $$
  Y_{ij}=X_i(T_{ij})+\varepsilon_{ij},
  \qquad j=1,\ldots,M_i.
  $$

- Assume
  $$
  E(\varepsilon_{ij})=0,\qquad
  \operatorname{Var}(\varepsilon_{ij})=\sigma_\varepsilon^2,
  $$

- Measurement error is independent of the latent process and design.
:::

::: {.content-hidden when-profile="short"}
## Backup: Standardized Products

- Define
  $$
  \widehat R_{ijk}
  =
  \frac{
  \{Y_{ij}-\widehat\mu(T_{ij})\}
  \{Y_{ik}-\widehat\mu(T_{ik})\}
  }{
  \widehat\sigma(T_{ij})\widehat\sigma(T_{ik})
  },
  \qquad j\neq k.
  $$

- Then fit candidate correlations to $\widehat R_{ijk}$.
:::

::: {.content-hidden when-profile="short"}
## Backup: Correlation Fit

- For family $\ell$ with parameter $\theta_\ell$:
  $$
  \widehat\theta_\ell
  =
  \arg\min_{\theta_\ell}
  \sum_{i}\sum_{j\neq k}
  w_{ijk}
  \left\{
  \widehat R_{ijk}
  -\rho_\ell(T_{ij},T_{ik};\theta_\ell)
  \right\}^2 .
  $$

- Weights can encode design, trimming, or numerical safeguards.
:::

::: {.content-hidden when-profile="short"}
## Backup: Simplex Averaging

- Let $\widehat R_\ell$ denote the fitted correlation matrix or grid surface for
  candidate family $\ell$.

- The averaged correlation is
  $$
  \widehat R_{\boldsymbol{\alpha}}
  =\sum_{\ell=1}^{L}\widehat\alpha_\ell\widehat R_\ell,
  \qquad
  \widehat{\boldsymbol{\alpha}}\in\Delta_L.
  $$
:::

::: {.content-hidden when-profile="short"}
## Backup: Boundary-Corrected Linear Span

- Linear correction starts from
  $$
  \{1,\;t,\;\sqrt{2}\cos(\pi t),\ldots,\sqrt{2}\cos(L\pi t)\}.
  $$

- Because $t'(0)=t'(1)=1$, this permits a common endpoint derivative.
:::

::: {.content-hidden when-profile="short"}
## Backup: Boundary-Corrected Quadratic Span

- Quadratic correction starts from
  $$
  \{1,\;t,\;t^2,\;\sqrt{2}\cos(\pi t),\ldots,\sqrt{2}\cos(L\pi t)\}.
  $$

- Since $t'(0)=1$, $t'(1)=1$, $2t|_{t=0}=0$, and $2t|_{t=1}=2$,
  the two endpoint derivatives can differ.
:::

::: {.content-hidden when-profile="short"}
## Backup: QR Construction

- Let $B$ be the evaluated design matrix for the finite span.

- Compute
  $$
  B = QR,
  $$

- The columns of $Q$ are orthonormal in the empirical inner product.

- The fitted values are obtained by projecting onto selected columns of $Q$.
:::

::: {.content-hidden when-profile="short"}
## Backup: Why Smooth Thresholding Still Fits

- The fixed orthonormal finite-span construction preserves the series-estimation
  logic used by the cosine estimator.

- For the Gram-Schmidt boundary basis, the same smoothed thresholding ideas can
  be used because the basis remains orthonormal after construction.

- Empirical QR variants require cross-validation because the orthonormalization
  is sample-design dependent.
:::

::: {.content-hidden when-profile="short"}
## Backup: No One Boundary Rule Wins Everywhere

- Endpoint correction is a bias-correction problem.

- If the true function has zero endpoint slopes, uncorrected cosine can be
  ideal.

- If the true function has nonzero endpoint slopes, uncorrected cosine can be
  structurally biased at the boundary.

- The applied answer is sensitivity analysis plus a defensible default.

:::

# References {.scrollable}

::: {#refs}
:::
